Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016

### 2019-04-03 draft-paper.Rmd
Minor wording changes only

---
title: "Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016"
author:
- familyname: Forbes
  othernames: Jeremy
  address: Monash University\newline Melbourne, Australia
  email: jeremy.forbes@monash.edu
  correspondingauthor: true
- familyname: Cook
  othernames: Di
  address: Monash University\newline Melbourne, Australia
- familyname: Hyndman
  othernames: Rob J
  address: Monash University\newline Melbourne, Australia
abstract: "This research examines the relationships between electoral socio-demographic characteristics and two-party preference in Australian federal elections, with the six elections from 2001 to 2016 considered. Socio-demographic information is derived from the Australian Census, which occurs every five years. Since a Census is not directly available for each election, a method of spatio-temporal imputation is employed to construct Census data for the electorates in 2004-2013 elections. This accounts for both spatial and temporal changes in electoral characteristics between Censuses. To capture any spatial heterogeneity, a spatial error model is estimated for each election, which incorporates a spatially structured random effect vector that can thought of as the unobserved political climate in each electorate. Over time, the impact of most socio-demographic characteristics that affect electoral two-party preference do not vary, with industry of work, incomes, household mobility and de facto relationships having strong effects in each of the six elections. Education and unemployment are amongst those that have varying effects. It is also found that between 2004 and 2013, the spatial effect is insignificant, meaning that electorates voted effectively independently. All data featured in this study has been contributed to the `eechidna` `R` package (available on CRAN)."
keywords: "federal election, Census, Australia, spatial modelling, imputation, socio-demographics, electorates, R, eechidna"
wpnumber:
jelcodes: C31, C33, D72
blind: false
cover: false
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
header-includes: \usepackage{float}
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

@@

Joining these two data sources is problematic as there in an inherent asynchronicity in the two types of events. A Census is conducted by the Australian Bureau of Statistics (ABS) every five years, whereas a federal election (conducted by the Australian Electoral Commission (AEC)) usually occurs every three years. The first problem addressed is that of constructing appropriate Census data for the 2004, 2007, 2010 and 2013 elections - election years in which a Census does not occur. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see @DavisStimson98, @Stimson06, @Liao09 and @Stimson09). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching with electoral data. To obtain appropriate Census data for these four elections, electoral socio-demographics are constructed using a spatio-temporal imputation that combines areal interpolation [@Goodchild1993] and linear time-interpolation. Collecting and wrangling the raw data, along with the imputation process, are detailed in Section \@ref(data). All data and associated documentation relating to this procedure are available in the `eechidna` `R` package [@eechidna], providing a resource for future analysis .

Previous work on modelling Australian federal elections have found that aggregate socio-demographics are relatively good predictors of voting outcomes. @Forrest01 use multiple regression to model the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998 as a function of Census variables from 1996. @Stimson06, @Stimson09 and @Stimson12 use principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. On the contrary, @Stimson09 models the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, finding that little of swing vote can be explained by Census data. Instead of analyzing a single election in isolation, this paper employs a consistent model framework across six elections so that temporal changes in the effects of socio-demographics can be observed, where each federal elections is modelled with a cross-sectional dataset. The use of a regression framework to examine these socio-political relationships over time is seemingly absent from previous Australian studies. It also appears that no study has attempted any type of statistical analysis of socio-demographics in conjunction with voter behaviour in Australia since 2007, making this paper distinctly different from those previous.

The cross-sectional dataset for each election used here consists of the two-party preferred vote (response variable), and a set of common socio-demographic variables (explanatory variables) that characterize each electorate. To obtain these datasets, socio-demographic variables are first standardized, and then a principal component analysis is used to group variables into "factors". To account for the inherent spatial structure of the data, a spatial error model is then fit for each election. These steps are discussed in Section \@ref(modelling). In Section \@ref(results) inference is conducted to see which effects are significant and how these effects change over time.

@@

Socio-demographic variables are derived from the Census of Population and Housing (Census), which is a survey of every household in Australia, recording information such as age, gender, ethnicity, education level and income. There have been four Censuses so far in the 21st century, conducted in 2001, 2006, 2011 and 2016. The Australian Bureau of Statistics (ABS) conducts the Census and publishes aggregated information. The ABS uses electoral boundaries as defined by the AEC at the time of each Census, which may not match those in place at the subsequent election. From the available Census information aggregated at the electorate level, 65 socio-demographic variables are defined for each of the electorates to be used in the analysis.

Raw data is sourced online from the AEC and ABS websites in .csv and .xlsx files. The formats of these files differ over the years, making extracting the appropriate information a big task. The functions available in the `dplyr` [@dplyr] and `readxl` [@readxl] `R` packages are particularly useful, as they provide fast consistent tools for data manipulation and functions to import .xlsx files (respectively). The 2001 and 2006 Census data are published in a format where the information for each electorate is held in a separate document making it difficult to use the `dplyr` tools. Instead, cells have to be selected from each individual file to construct the desired variables. All scripts required for the data wrangling process can be found in the github repository for the `eechidna` `R` package [@eechidna], along with the raw data. The `eechidna` package makes this study entirely reproducible and provides a resource to help wrangle data for future Censuses and elections, when they become available.

@@

Between 2001 and 2016 there were six elections and four Censuses (see Figure \@ref(fig:timeline)). Electoral boundaries are redistributed regularly by the AEC, meaning that only in the years where both a Census and election occur will the boundaries match - the case for the 2001 and 2016 elections. Therefore, for the four elections between 2004 and 2013, both temporal and spatial differences in electorates need to be accounted for when joining the electoral two-party preferred vote with Census data. For these elections a spatio-temporal imputation method is employed to obtain electoral socio-demographics. This method uses Census information from both before and after the election of interest.

@@

To account for spatial differences, the piece-wise approximation method in @Goodchild1993 is adopted. Consider a map of source zones $s = 1,...,S$, for which socio-demographic information is available, and a set of target zones $t = 1,...,T$ for which information is to be imputed. This is described in the context of a single election, and a single neighbouring Census.

Let the map of electoral boundaries at the time of a Census define the source zones, and let the boundaries at the time of the election be the target zones. Denote the area of intersection between a source zone $s$ and a target zone $t$ as $A_{s,t}$. Additionally, let the population of the source zone $s$ be $U_s$ and the population of intersection between source zone $s$ and target zone $t$ be $P_{s,t}$. The estimated population of intersection is given by:

$$\hat{P}_{s,t} = \frac{U_s*A_{s,t}}{\sum_{t=1}^T A_{s,t}}, \mbox{ for all } s=1,...,S \mbox{ and } t=1,...,T.$$
Note that this estimator implicitly assumes that populations are uniformly distributed within each source zone.

In order to calculate socio-demographic information for each of the target zones, a weighted average is taken using the estimated intersection populations as weights. Denote a given Census variable for the target zone by $C_t$, and the same Census variable for the source zone as $D_s$. Then, estimate $C_t$ using:

$$\hat{C}_t = \frac{\sum_{s=1}^{S}{D_s*\hat{P}_{s,t}}}{\sum_{s=1}^{S}{\hat{P}_{s,t}}}, \mbox{ for each } t=1,...,T.$$
This concludes the spatial imputation of the socio-demographic characteristics for one target zone (a single electoral boundary), at time of only one of the neighbouring Censuses. This process is repeated for all of the target zones, and then for the other neighbouring Census.

To account for temporal changes, linear interpolation is used between Census years to get the final estimate of a Census variable for the target zone in the election year. Let $y_1$ be the year of the Census preceding an election, let $y_2$ be the year of the election and $y_3$ be the year of the Census that follows. Add this year subscript to the Census variable estimate $\hat{C}_t$, resulting in $\hat{C}_{t,y}$. Linear interpolating between these Census years results an imputed value for the election year, given by

$$\hat{C}_{t,y_2} = \frac{y_3-y_2}{y_3-y_1}*\hat{C}_{t,y_1} + \frac{y_2-y_1}{y_3-y_1}*\hat{C}_{t,y_3}.$$
Implicitly this assumes that population characteristics change in a linear manner over time.

@@

Census data is publicly available that different levels of aggregation, ranging from SA1 (over 50,000 zones) to electoral divisions (150 zones). For this study electoral divisions are used as source zones, and the imputation method is applied to produce socio-demographic variables for each of the 2004, 2007, 2010 and 2013 elections. As mentioned earlier, there is no need to impute socio-demographic variables for the 2001 and 2016 elections. To illustrate the method, consider the imputation of socio-demographic variables for the electorate of Hume in New South Wales (NSW) at the time of the 2013 federal election. The boundaries shown in Figure \@ref(fig:hume13) define all target zones in NSW for 2013, with the target zone of interest (Hume) shaded purple.

@@

Source Zone (2016) | Proportion | Source Zone Population | Population Allocated to Purple Region: $\hat{P}_{s,t}$
--------------------|----------|-----------|---------------------------------------------
HUME              | 0.9654     | 150643                   | 145427
RIVERINA          | 0.2511     | 155793                   | 39117
EDEN-MONARO       | 0.1109     | 147532                   | 16358
CANBERRA          | 0.028      | 196037                   | 548
FENNER            | 0.0023     | 202955                   | 474
WHITLAM           | 0.0006     | 152280                   | 92
GILMORE           | 0.0006     | 150436                   | 86
CALARE            | 0.0001     | 161298                   | 21

@@

Now consider the socio-demographic variable $AusCitizen$ - the proportion of people in the region who are Australian citizens.

@@

|Source Zone (2016)   | AusCitizen (%): $D_s$| Estimated Population Allocated to Purple Region: $\hat{P}_{s,t}$|
|--------------------------|-------------------|------------------------------------------------|
|HUME            |              90.02|                                          145427|
|RIVERINA        |              89.11|                                           39117|
|EDEN-MONARO     |              88.00|                                           16358|
|CANBERRA        |              85.48|                                             548|
|FENNER          |              83.64|                                             474|
|WHITLAM         |              89.52|                                              92|
|GILMORE         |              89.03|                                              86|
|CALARE          |              87.56|                                              21|

@@

### Industry and type of work
Electorates with higher proportions of workers in mining, gas, water, agriculture, waste and electricity (grouped as `Extractive` industries) are consistently linked with higher support for the Liberal party, with the magnitude of this effect slightly increasing over the years (see row 3 in Figure \@ref(fig:partresplot1)). This is unsurprising, as the Liberal party has close ties with these traditional energy industries, and typically present policies to reduce taxation on energy production. Furthermore, electorates with more workers in construction or manufacturing industries (`Transformative`) are also more likely to support the Liberal party (see row 4 in Figure \@ref(fig:partresplot1)).

>>>>

---
title: "Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016"
author:
- familyname: Forbes
  othernames: Jeremy
  address: Monash University\newline Melbourne, Australia
  email: jeremy.forbes@monash.edu
  correspondingauthor: true
- familyname: Cook
  othernames: Di
  address: Monash University\newline Melbourne, Australia
- familyname: Hyndman
  othernames: Rob J
  address: Monash University\newline Melbourne, Australia
abstract: "We examine the relationships between electoral socio-demographic characteristics and two-party preference in the six Australian federal elections held between 2001 to 2016. Socio-demographic information was derived from the Australian Census, which occurs every five years. Since a Census is not directly available for each election, spatio-temporal imputation was employed to estimate Census data for the electorates at the time of each election. This accounts for both spatial and temporal changes in electoral characteristics between Censuses. To capture any spatial heterogeneity, a spatial error model was estimated for each election, which incorporates a spatially structured random effect vector that can be thought of as the unobserved political climate in each electorate. Over time, the impact of most socio-demographic characteristics that affect electoral two-party preference do not vary, with industry of work, incomes, household mobility and de facto relationships having strong effects in each of the six elections. Education and unemployment are amongst those that have varying effects. It is also found that between 2004 and 2013, the spatial effect was insignificant, meaning that electorates voted effectively independently. All data featured in this study has been contributed to the `eechidna` `R` package (available on CRAN)."
keywords: "federal election, Census, Australia, spatial modelling, imputation, socio-demographics, electorates, R, eechidna"
wpnumber:
jelcodes: C31, C33, D72
blind: false
cover: false
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
header-includes: \usepackage{float}
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

@@

Joining these two data sources is problematic as there in an inherent asynchronicity in the two types of events. A Census is conducted by the Australian Bureau of Statistics (ABS) every five years, whereas federal elections, conducted by the Australian Electoral Commission (AEC), usually occur every three years or so. The first problem addressed is that of constructing appropriate Census data for the 2004, 2007, 2010 and 2013 elections --- election years in which a Census does not occur. The predominant approach in previous studies was to join voting outcomes to the nearest Census, without accounting for any temporal differences (see @DavisStimson98, @Stimson06, @Liao09 and @Stimson09). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching with electoral data. To obtain appropriate "Census-like" data for these four elections, electoral socio-demographics are constructed using a spatio-temporal imputation that combines areal interpolation [@Goodchild1993] and linear time-interpolation. Collecting and wrangling the raw data, along with the imputation process, are detailed in Section \@ref(data). All data and associated documentation relating to this procedure are available in the `eechidna` `R` package [@eechidna], providing a resource for future analysis.

Previous work on modelling Australian federal elections has found that aggregate socio-demographics are relatively good predictors of voting outcomes. @Forrest01 used multiple regression to model the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998 as a function of Census variables from 1996. @Stimson06, @Stimson09 and @Stimson12 used principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. In contrast, @Stimson09 models the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, finding that little of the swing vote can be explained by Census data. Instead of analyzing a single election in isolation, this paper employs a consistent model framework across six elections so that temporal changes in the effects of socio-demographics can be observed, where each federal elections is modelled with a cross-sectional dataset. The use of a regression framework to examine these socio-political relationships over time is seemingly absent from previous Australian studies. It also appears that no study has attempted any type of statistical analysis of socio-demographics in conjunction with voter behaviour in Australia since 2007, making this paper distinctly different from previous work.

The cross-sectional dataset for each election used here consists of the two-party preferred vote (as the response variable), and a set of common socio-demographic variables (as the explanatory variables) that characterize each electorate. To prepare these datasets, socio-demographic variables are first standardized, and then a principal component analysis is used to group variables into "factors". To account for the inherent spatial structure of the data, a spatial error model is then estimated for each election. These steps are discussed in Section \@ref(modelling). In Section \@ref(results) inference is conducted to see which effects are significant and how these effects change over time.

@@

Socio-demographic variables were derived from the Australian Census of Population and Housing (Census), which is a survey of every household in Australia, recording information such as age, gender, ethnicity, education level and income. There have been four Censuses so far in the 21st century, conducted in 2001, 2006, 2011 and 2016. The Australian Bureau of Statistics (ABS) conducts the Census and publishes aggregated information. The ABS uses electoral boundaries as defined by the AEC at the time of each Census, which may not match those in place at the subsequent and previous elections. From the available Census information aggregated at the electorate level, 65 socio-demographic variables were defined for each of the electorates to be used in the analysis.

Raw data was sourced online from the AEC and ABS websites in `.csv` and `.xlsx` files. The formats of these files differ over the years, making extracting the appropriate information a big task. The functions available in the `dplyr` [@dplyr] and `readxl` [@readxl] `R` packages are particularly useful, as they provide fast consistent tools for data manipulation and functions to import `.xlsx` files (respectively). The 2001 and 2006 Census data are published in a format where the information for each electorate is held in a separate document making it difficult to use the `dplyr` tools. Instead, cells have to be selected from each individual file to construct the desired variables. All scripts required for the data wrangling process can be found in the github repository for the `eechidna` `R` package [@eechidna], along with the raw data. The `eechidna` package makes this study entirely reproducible and provides a resource to help wrangle data for future Censuses and elections, when they become available.

@@

Between 2001 and 2016 there were six elections and four Censuses (see Figure \@ref(fig:timeline)). Electoral boundaries are redistributed regularly by the AEC, meaning that only in the years where both a Census and election occur are all boundaries likely to match --- the case for the 2001 and 2016 elections. Therefore, for the four elections between 2004 and 2013, both temporal and spatial differences in electorates need to be accounted for when joining the electoral two-party preferred vote with Census data. For these elections a spatio-temporal imputation method was employed to obtain electoral socio-demographics. This method uses Census information from both before and after the election of interest.

@@

To account for spatial differences, the piece-wise approximation method in @Goodchild1993 is adopted. Consider a map of source zones $s = 1,\dots,S$, for which socio-demographic information is available, and a set of target zones $t = 1,\dots,T$ for which information is to be imputed. This is described in the context of a single election, and a single neighbouring Census.

Let the map of electoral boundaries at the time of a Census define the source zones, and let the boundaries at the time of the election be the target zones. Denote the area of intersection between a source zone $s$ and a target zone $t$ as $A_{s,t}$. Additionally, let the population of the source zone $s$ be $U_s$ and the population of intersection between source zone $s$ and target zone $t$ be $P_{s,t}$. The estimated population of intersection is given by
$$
  \hat{P}_{s,t} = \frac{U_s*A_{s,t}}{\sum_{t=1}^T A_{s,t}},
  \quad\text{for all $s=1,\dots,S$ and $t=1,\dots,T$}.
$$
Note that this estimator implicitly assumes that populations are uniformly distributed within each source zone.

In order to calculate socio-demographic information for each of the target zones, a weighted average is taken using the estimated intersection populations as weights. Denote a given Census variable for the target zone by $C_t$, and the same Census variable for the source zone as $D_s$. Then, estimate $C_t$ using
$$
  \hat{C}_t = \frac{\sum_{s=1}^{S}{D_s*\hat{P}_{s,t}}}%
  {\sum_{s=1}^{S}{\hat{P}_{s,t}}},
  \quad\text{for each $t=1,\dots,T$}.
$$
This concludes the spatial imputation of the socio-demographic characteristics for one target zone (a single electoral boundary), at the time of only one of the neighbouring Censuses. This process is repeated for all of the target zones, and then for the other neighbouring Census.

To account for temporal changes, linear interpolation is used between Census years to get the final estimate of a Census variable for the target zone in the election year. Let $y_1$ be the year of the Census preceding an election, let $y_2$ be the year of the election, and $y_3$ be the year of the Census that follows. Add this year subscript to the Census variable estimate $\hat{C}_t$, resulting in $\hat{C}_{t,y}$. Linear interpolating between these Census years results an imputed value for the election year, given by
$$
  \hat{C}_{t,y_2} = \frac{y_3-y_2}{y_3-y_1} * \hat{C}_{t,y_1} +
    \frac{y_2-y_1}{y_3-y_1} * \hat{C}_{t,y_3}.
$$
Implicitly this assumes that population characteristics change in a linear manner over time.

@@

Census data is publicly available at different levels of aggregation, ranging from SA1 (over 50,000 zones) to electoral divisions (150 zones). For this study, electoral divisions are used as source zones, and the imputation method is applied to produce socio-demographic variables for each of the 2004, 2007, 2010 and 2013 elections. As mentioned earlier, there is no need to impute socio-demographic variables for the 2001 and 2016 elections. To illustrate the method, consider the imputation of socio-demographic variables for the electorate of Hume in New South Wales (NSW) at the time of the 2013 federal election. The boundaries shown in Figure \@ref(fig:hume13) define all target zones in NSW for 2013, with the target zone of interest (Hume) shaded purple.

@@

Source Zone (2016) | Proportion | Source Zone Population | Population Allocated to Purple Region: $\hat{P}_{s,t}$
--------------------|----------|-----------|--------------------------------------------:
HUME              | 0.9654     | 150643                   | 145427
RIVERINA          | 0.2511     | 155793                   | 39117
EDEN-MONARO       | 0.1109     | 147532                   | 16358
CANBERRA          | 0.0280     | 196037                   | 548
FENNER            | 0.0023     | 202955                   | 474
WHITLAM           | 0.0006     | 152280                   | 92
GILMORE           | 0.0006     | 150436                   | 86
CALARE            | 0.0001     | 161298                   | 21

@@

Now consider the socio-demographic variable *AusCitizen*, the proportion of people in the region who are Australian citizens.

@@

|Source Zone (2016)   | AusCitizen (%): $D_s$| Estimated Population Allocated to Purple Region: $\hat{P}_{s,t}$|
|--------------------------|-------------------|-----------------------------------------------:|
|HUME            |              90.02|                                          145427|
|RIVERINA        |              89.11|                                           39117|
|EDEN-MONARO     |              88.00|                                           16358|
|CANBERRA        |              85.48|                                             548|
|FENNER          |              83.64|                                             474|
|WHITLAM         |              89.52|                                              92|
|GILMORE         |              89.03|                                              86|
|CALARE          |              87.56|                                              21|

@@

### Industry and type of work

Electorates with higher proportions of workers in mining, gas, water, agriculture, waste and electricity (grouped as `Extractive` industries) are consistently linked with higher support for the Liberal party, with the magnitude of this effect slightly increasing over the years (see row 3 in Figure \@ref(fig:partresplot1)). This is unsurprising, as the Liberal party has close ties with these traditional energy industries, and typically present policies to reduce taxation on energy production. Furthermore, electorates with more workers in construction or manufacturing industries (`Transformative`) are also more likely to support the Liberal party (see row 4 in Figure \@ref(fig:partresplot1)).
